---
title: 'Lab 1: Question 2'
author: "Jeffrey Adams, Brittany Dougall, Jerico Johns, Li Jin"
output: pdf_document
---

```{r load packages, echo=TRUE, warning=FALSE, message=FALSE}
install.packages("rstatix", repos = "http://cran.us.r-project.org")
install.packages("coin", repos = "http://cran.us.r-project.org")
library(coin)
library(rstatix)
library(dplyr)
library(ggplot2) 
library(tidyverse)
```

```{r load and clean data, echo=FALSE, warning=TRUE, message=FALSE}
# Load the data
Data <- haven::read_dta(file = 'anes_timeseries_2020_stata_20210211.dta')

# Create a new data frame with only the columns of interest
df <- Data[c("V201018", "V201228", "V201151", "V201153")]

# Rename the data frame columns using meaningful labels
df <- df %>%
  rename(registered_party = V201018, 
         party_affiliation = V201228,
         biden_rating = V201151,
         harris_rating = V201153,
         )

# Filter the data frame to only include rows where a valid enthusiasm score was entered for both Biden and Harris
valid_scores <- subset(df,df$biden_rating>=0 & df$biden_rating<=100 & df$harris_rating >= 0 & df$harris_rating <= 100)

# Filter the data frame from the above step to include only those who consider themselves to be affiliated with the Democratic party
democratic_rating_df <- valid_scores[(valid_scores$party_affiliation == 1), ]

# Compute Biden's score minus Harris' score for each respondent
democratic_rating_df <- democratic_rating_df %>% mutate(difference = biden_rating - harris_rating)
```
## Importance and Context
In presidential elections, presidential and vice presidential candidates run as a joint ticket. A significant difference in public enthusiasm between members of a joint ticket may influence voting patterns. For example, a strong favorable opinion regarding one of the members of a ticket may motivate a voter to still vote for the ticket even if the voter feels ambivalent about the other ticket member. Say, if a voter feels ambivalent about Harris but strongly about Biden, the presence of Biden on the ticket may motivate the voter to still vote for this joint ticket. Conversely, if enthusiasm about one candidate is low, the presence of the other member may not be sufficient to motivate the voter to choose the ticket. For example, if a voter had little enthusiasm for Biden, the presence of Harris on the ticket may not be sufficient to inspire the voter to choose the Democratic ticket. In this scenario, voters who would otherwise vote Democrat may choose to vote for another candidate ticket or to abstain from voting at all. For this reason, it could be useful to determine if Democrats are more enthusiastic about Biden or Harris.

## Description of Data
Data is collected from an ANES 2020 time series study using a cross-sectional random sample to survey individuals regarding their political opinions and voting behavior in the U.S. presidential election. To operationalize our research question of whether or not Democratic voters are more enthusiastic about Biden or Harris, we first compared the number of respondents who are registered Democrats from field V201018 to the number who consider themselves to be Democrat from field V201228. Both fields contain numeric scores that correspond to a political party or a number that corresponds to a reason for the absence of a political party entry. As seen below, the number of respondents who are registered Democrats (first result) is less than the number of respondents who consider themselves to be Democrat (the second result). For this reason, we selected the party affiliation field (V201228) to identify Democrat respondents (filtering the field to select respondents with an affiliation value of $1$).

```{r, echo=FALSE, warning=TRUE, message=FALSE}
# Count the number voters who are registered Democrats versus the number who consider themselves Democrat
NROW(subset(df, df$registered_party == 1))
NROW(subset(df, df$party_affiliation == 1))
```
V201151 contains voter enthusiasm scores for Biden and V201153 contains voter enthusiasm scores for Harris. Both fields contain numeric scores ranging from $0 - 100$ to indicate voter enthusiasm level, as well as the following values for both candidates: $998$ (don't know how to rate), $-4$ (technical error), $-9$ (refused). V201153 contains the additional potential entry for Harris: $999$ (don't know the candidate). We filtered the data frame to only included rows where candidates had a valid score. The first result below represents the number of rows where a valid score was entered for Biden but not Harris, the second result the number of rows with a valid score for Harris but not Biden, and the third result the number of rows with a valid score for both candidates. Since the majority of respondents entered a score for both Biden and Harris, we only included rows where a valid enthusiasm score between $0$ and $100$ was entered for both candidates.

```{r, echo=FALSE, warning=TRUE, message=FALSE}
# Checking the number of rows for respondents who consider themselves Democrats where Biden has a valid enthusiasm score but Harris does not and vice versa reveals that the majority of respondents enter a valid score for both candidates if a valid score is entered
  # For this reason, we will use a paired test
NROW(subset(df,df$party_affiliation == 1 & df$biden_rating >= 0 & df$biden_rating <= 100 & (df$harris_rating %in% c(999, 998, -4, -9))))
NROW(subset(df,df$party_affiliation == 1 & df$harris_rating >= 0 & df$harris_rating <= 100 & (df$biden_rating %in% c(999, 998, -4, -9))))
NROW(democratic_rating_df)
```

## Most appropriate test
As seen above, the majority of Democrats entered a score for both candidates, indicating that we should choose a paired test. The data for our scale is ordinal numeric (0 - 100), where one cannot reasonably interpret meaning behind a single unit. However, since we are interested in testing if the mean enthusiasm score is higher for Biden or Harris among Democratic voters, and are taking the difference between scores across each respondent, we are imposing a metric scale. As seen in the below histogram of Biden enthusiasm scores minus Harris enthusiasm scores, the differences in enthusiasm scores do not appear to be normally distributed. The following qqplot of the correlation between this difference in scores with the norm also suggests that the data do not follow a normal distribution (meaning we should choose a non-parametric test). Therefore, we chose to conduct a Wilcoxon signed-rank test to test our null hypothesis that the difference in enthusiasm scores between Biden and Harris for a given Democratic voter is $0$ (selecting a two sided test to be more conservative in rejecting the null hypothesis).

```{r fig.width = 7, fig.height= 3.5, echo=FALSE, warning=TRUE, message=FALSE}

# Using a histogram plot the differences in scores suggests that the difference in scores is not normally distributed
ggplot(democratic_rating_df, aes(x = difference)) + 
  geom_histogram(fill = "blue", binwidth = 10) +
  scale_x_continuous(breaks = seq(-100, 100, 10)) +
  labs(title = "Differences in Enthusiasm Scores",
       x = "Biden Score - Harris Score",
       y = "Count") + 
  theme_classic()
```
``````{r fig.width = 7, fig.height= 3.5, echo=FALSE, warning=TRUE, message=FALSE}

# Plotting the correlation between the difference (Biden score minus Harris score) and the norm also suggests deviations from normality
qqnorm(democratic_rating_df$difference, pch = 1, frame = FALSE)
qqline(democratic_rating_df$difference, col = "steelblue", lwd = 2)
```
The assumptions behind this test are: $1$) the data have a metric scale, $2$) I.I.D. data - each pair of measurements is drawn from the same distribution, independently of all other pairs, and $3$) the distribution of the difference between enthusiasm scores is symmetric around some mean. Per the documentation for the columns for Biden and Harris' enthusiasm scores, valid scores for both columns take on values between $0$ and $100$ and we impose a metric scale by taking the difference in these values, meeting the metric scale requirement. Scores outside of this range will be excluded from our analysis. Per the documentation, only $1$ member of each household was surveyed and respondents were randomly selected from the USPS delivery sequence file. It is possible that individuals who chose to participate encouraged others outside their household to participate; however, given that all addresses within the $50$ states and D.C. were given equal likelihood of being selected, this possibility is rare (meeting the I.I.D. requirement). From the above histogram distribution of Biden enthusiasm score minus Harris enthusiasm score, the distribution of the difference in scores is approximately symmetric around the median, meeting the third requirement. 

## Test, results and interpretation
```{r, echo=FALSE, warning=TRUE, message=FALSE}
# Use the Wilcox Signed Rank test to test the null hypothesis that there is no significant difference in public enthusiasm for Biden and Harris
wilcox.test(democratic_rating_df$biden_rating, democratic_rating_df$harris_rating, alternative = "two.sided", paired = TRUE)

democratic_rating_df.long <- democratic_rating_df %>%
  gather(key = "group", value = "enthusiasm_score", biden_rating, harris_rating)

democratic_rating_df.long %>% 
  wilcox_effsize(enthusiasm_score ~ group, paired = TRUE)
```
Since the p-value ($0.0001215$) is less than our significance level of 0.05, we should reject the null hypothesis that there is no difference in mean enthusiasm for Biden and Harris among Democratic voters. However, while the difference in Biden and Harris' mean enthusiasm scores is statistically significant, the effect size is small. This suggests that the difference in mean enthusiasm has little practical significance.